\artus -- Framework for Event-based Data Analysis in High Energy Physics

### 2015-09-09 Core/doc/latex/paper.tex
paper: added some authors
paper: 'Artus' in \textsc

@@

\begin{frontmatter}
\title{Artus -- Framework for Event-based Data Analysis in High Energy Physics}

\author[kit]{Thomas M\"uller}
\address[kit]{Karlsruhe Institute of Technology, Wolfgang-Gaede-Stra\ss{}e 1, D-76131 Karlsruhe}

\begin{abstract}
We present Artus, a data-processing application for high energy physics experiments. Artus is designed for large-scale data analysis in a collaborative environment.
The architecture design choices take into account typical challenges and are based on experiences with similar applications.
The structure of the framework and its advantages are described.
An example usecase and performance measurements are presented.
The framwork is well-tested and successfully used by several analysis groups.

@@

\section{The Artus Framework \label{section_artus_motivation}}

@@

Following the structure of the CMSSW framework, the Artus framework\footnote{\url{https://github.com/artus-analysis/Artus}} provides a front-to-end solution for the analysis of event-based data. The basic concept of modules processing the events within an event loop is complemented by tools for the configuration of every step. The complete structure is explained in the next section~\ref{section_artus_structure}.

This general framework is applicable to any kind of event processing analysis. The well-tested core is shared among various analyses, providing a reliable fundament for any new analysis. The highly modular structure guides the user to develop analysis software that is easily maintainable and constructively extensible. Thus, thes Artus framework avoids the aforementioned problems of separate user code for individual analyses.

The Artus framework is written in C++. The core of the framework only depends on the boost library\footnote{\url{http://www.boost.org/}}. Further dependencies may be introduced by the actual analysis code. C++ is chosen for performance reasons and in order to be able to integrate existing frameworks and libraries that are commonly used in high energy physics.

@@

\begin{figure}[p]
\centering \input{figures/artus_structure}
\caption[Structure of an Artus analysis.]{Structure of an Artus analysis. The input is read by an event provider. Within the pipelines the event content is analysed by the processors. Consumers in local pipelines write results to a common output. All parts of the analysis are configurable.}
\label{figure_artus_structure}

@@

In the Artus framework each concrete analysis is defined by a set of object types. It needs an implementation of the classes for settings, the event and the product. The settings class defines the tags that should be able to be read from the configuration file and the corresponding types of the values. The event class defines the event content being read from the input and the product class contains members for new quantities being calculated by the analysis. These three classes need to be derived from the base classes in the core of the framework.

An implementation of an event provider must be available. The event provider must know how to read event content from the input files and how to store them in the event object. Furthermore, a factory managing the creation of analysis-specific processors needs to exists. These classes also have to be derived from the Artus base classes.

The most important part of an analysis are the processors. The abstract processors of the Artus core need to be implemented to realise the physics needs of the analysis in the form of code. New processors have to be registered in the factory. After the basic structure of an analysis is set up, most of changes and additions are only related to the processors.

@@

The structure of the framework is illustrated in figure~\ref{figure_artus_harry_plotter}. It follows the idea of processors in the Artus framework. Three kinds of processor modules are executed subsequently:
\begin{enumerate}
\item One or, in special cases, more input modules are run to read in data from an input source. The input format is defined by the input module. The most commonly used input module is the one that reads histograms or other objects from ROOT files.
\item An arbitrary number of analysis modules is appended. They can perform calculations based on the inputs or the results of previously run analysis modules.
\item One or, in special cases, multiple plot modules are executed in the last step. Their function is to output the processed data in the specified format. In most of the cases, the output is a plot but it is also possible to store the objects in new ROOT files.
\end{enumerate}
All processors share a common plot data object and possible meta-data where they can add new information and read the existing one. For example, the input modules add histograms read in from input files and the plot modules access these histograms.

@@

Currently, the Artus framework is used for various user analysis of CMS data ranging from jet energy calibration and QCD studies to analyses in the scope of Higgs boson measurements and searches for physics beyond the Standard Model. As an example, the technical aspects of the $\PH\to\Pgt\Pgt$ analysis presented in this thesis are explained in the following.

The $\PH\to\Pgt\Pgt$ analysis developed in the scope of this thesis makes use of skimmed data sets in the Kappa format. The Kappa framework\footnote{\url{https://github.com/KappaAnalysis/Kappa}} is designed for skimming CMS data into small n-tuples that can be analysed independently of the CMSSW software. The only dependence is the ROOT framework. The n-tuples store high-level information such as four-momentum vectors, collections of leptons and vertices. Similar as the Artus framework, Kappa aims to satisfy general analysis needs.

A large part of the Artus software is specialised on the analysis of Kappa n-tuples\footnote{\url{https://github.com/artus-analysis/Artus/tree/master/KappaAnalysis}}. An event provider is available for reading Kappa n-tuples. The event class contains members for all important physics objects used in analyses. The branches of the n-tuples, that should be read in from the input files, are configurable. Reducing the amount of information transferred from the input files accelerates the analysis significantly.

@@

Table~\ref{table_artus_runtime_comparison} shows a runtime measurement for the baseline part of the $\PH\to\Pgt\Pgt$ in the four main channels ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$ and $\Pgt\Pgt$). It illustrates the main advantage of the Artus structure: sub-analyses with different sets of configurations can be performed in the same run avoiding the multiple IO overhead when the configurations would be performed sequentially in different runs.

>>>>

\newcommand{\software}[1]{\textsc{#1}\xspace}
\newcommand{\artus}{\software{Artus}}

@@

\begin{frontmatter}
\title{\artus -- Framework for Event-based Data Analysis in High Energy Physics}

\author[kit]{Thomas M\"uller, Thomas Hauth, Dominik Haitz}
\address[kit]{Karlsruhe Institute of Technology, Wolfgang-Gaede-Stra\ss{}e 1, D-76131 Karlsruhe}

\begin{abstract}
We present \artus, a data-processing framework for high energy physics experiments. \artus is designed for large-scale data analysis in a collaborative environment.
The architecture design choices take into account typical challenges and are based on experiences with similar applications.
The structure of the framework and its advantages are described.
An example usecase and performance measurements are presented.
The framework is well-tested and successfully used by several analysis groups.

@@

\section{The \artus Framework \label{section_artus_motivation}}

@@

Following the structure of the CMSSW framework, the \artus framework\footnote{\url{https://github.com/artus-analysis/\artus}} provides a front-to-end solution for the analysis of event-based data. The basic concept of modules processing the events within an event loop is complemented by tools for the configuration of every step. The complete structure is explained in the next section~\ref{section_artus_structure}.

This general framework is applicable to any kind of event processing analysis. The well-tested core is shared among various analyses, providing a reliable fundament for any new analysis. The highly modular structure guides the user to develop analysis software that is easily maintainable and constructively extensible. Thus, thes \artus framework avoids the aforementioned problems of separate user code for individual analyses.

The \artus framework is written in C++. The core of the framework only depends on the boost library\footnote{\url{http://www.boost.org/}}. Further dependencies may be introduced by the actual analysis code. C++ is chosen for performance reasons and in order to be able to integrate existing frameworks and libraries that are commonly used in high energy physics.

@@

\begin{figure}[p]
\centering \input{figures/artus_structure}
\caption[Structure of an \artus analysis.]{Structure of an \artus analysis. The input is read by an event provider. Within the pipelines the event content is analysed by the processors. Consumers in local pipelines write results to a common output. All parts of the analysis are configurable.}
\label{figure_artus_structure}

@@

In the \artus framework each concrete analysis is defined by a set of object types. It needs an implementation of the classes for settings, the event and the product. The settings class defines the tags that should be able to be read from the configuration file and the corresponding types of the values. The event class defines the event content being read from the input and the product class contains members for new quantities being calculated by the analysis. These three classes need to be derived from the base classes in the core of the framework.

An implementation of an event provider must be available. The event provider must know how to read event content from the input files and how to store them in the event object. Furthermore, a factory managing the creation of analysis-specific processors needs to exists. These classes also have to be derived from the \artus base classes.

The most important part of an analysis are the processors. The abstract processors of the \artus core need to be implemented to realise the physics needs of the analysis in the form of code. New processors have to be registered in the factory. After the basic structure of an analysis is set up, most of changes and additions are only related to the processors.

@@

The structure of the framework is illustrated in figure~\ref{figure_artus_harry_plotter}. It follows the idea of processors in the \artus framework. Three kinds of processor modules are executed subsequently:
\begin{enumerate}
\item One or, in special cases, more input modules are run to read in data from an input source. The input format is defined by the input module. The most commonly used input module is the one that reads histograms or other objects from ROOT files.
\item An arbitrary number of analysis modules is appended. They can perform calculations based on the inputs or the results of previously run analysis modules.
\item One or, in special cases, multiple plot modules are executed in the last step. Their function is to output the processed data in the specified format. In most of the cases, the output is a plot but it is also possible to store the objects in new ROOT files.
\end{enumerate}
All processors share a common plot data object and possible meta-data where they can add new information and read the existing one. For example, the input modules add histograms read in from input files and the plot modules access these histograms.

@@

Currently, the \artus framework is used for various user analysis of CMS data ranging from jet energy calibration and QCD studies to analyses in the scope of Higgs boson measurements and searches for physics beyond the Standard Model. As an example, the technical aspects of the $\PH\to\Pgt\Pgt$ analysis presented in this thesis are explained in the following.

The $\PH\to\Pgt\Pgt$ analysis developed in the scope of this thesis makes use of skimmed data sets in the Kappa format. The Kappa framework\footnote{\url{https://github.com/KappaAnalysis/Kappa}} is designed for skimming CMS data into small n-tuples that can be analysed independently of the CMSSW software. The only dependence is the ROOT framework. The n-tuples store high-level information such as four-momentum vectors, collections of leptons and vertices. Similar as the \artus framework, Kappa aims to satisfy general analysis needs.

A large part of the \artus software is specialised on the analysis of Kappa n-tuples\footnote{\url{https://github.com/artus-analysis/Artus/tree/master/KappaAnalysis}}. An event provider is available for reading Kappa n-tuples. The event class contains members for all important physics objects used in analyses. The branches of the n-tuples, that should be read in from the input files, are configurable. Reducing the amount of information transferred from the input files accelerates the analysis significantly.

@@

Table~\ref{table_artus_runtime_comparison} shows a runtime measurement for the baseline part of the $\PH\to\Pgt\Pgt$ in the four main channels ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$ and $\Pgt\Pgt$). It illustrates the main advantage of the \artus structure: sub-analyses with different sets of configurations can be performed in the same run avoiding the multiple IO overhead when the configurations would be performed sequentially in different runs.
